Go Implementation of Up-To Techniques for Equivalence of Weighted Languages

### 2020-09-30 paper/paper.tex
preliminary definitions

\theoremstyle{plain}% default
\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newtheorem*{cor}{Corollary}
\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{conj}{Conjecture}[section]
\newtheorem{exmp}{Example}[section]
\newtheorem{exrc}[exmp]{Exercise}
\theoremstyle{remark}
\newtheorem*{comm}{Comment}
\newtheorem*{note}{Note}
\newtheorem{caso}{Case}

@@

\newcommand{\diag}[1]{\text{diag}\left(#1\right)}
\newcommand{\mspan}[1]{\text{span}\left(#1\right)}
\newcommand{\dom}{\text{dom}}
\newcommand{\mker}[1]{\text{ker}\left(#1\right)}
\newcommand{\mrank}[1]{\text{rank}\left(#1\right)}

@@

\author{
  Alessandro Cheli \\
  Undergraduate Student \\
  Department of Computer Science \\
  Università di Pisa\\
  Pisa, PI 56127 \\
  \texttt{a.cheli6@studenti.unipi.it} \\
  %% examples of more authors
  %% \And
  %% Elias D.~Striatum \\
  %% Department of Electrical Engineering\\
  %% Mount-Sheikh University\\
  %% Santa Narimana, Levand \\
  %% \texttt{stariate@ee.mount-sheikh.edu} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

@@

In \cite{DBLP:journals/corr/Bonchi0K17}, up-to techniques are defined for
weighted systems over arbitrary semirings, while in \cite{BONCHI201277}, up-to techniques
are defined for Linear Weighted Automata (LWAs), under a more abstract coalgebraic perspective.
For ease of implementation, we focus only on weighted automata over the field of real numbers 
$\R$.

@@

\begin{defn}
  A \textit{weighted automaton} over a field $\K$ and an alphabet $A$ is a triple 
  $(X,o,t)$ such that $X$ is a finite set of states, $t = (t_a : X \to K)_{a \in A}$
  is a set of transition functions indexed over the symbols of the alphabet $A$ and $o : X 
  \to \K$ is the output function. The transition functions $t_a$ are represented as
  $\K^{X \times X}$ matrices. $o \in \K^{1 \times X}$ is represented as a row vector.
  $t_a(v)$ denotes the vector obtained by multiplying the matrix $t_a$ by the column 
  vector $v  \in \K^{X \times 1}$. $o(v)$ denotes the scalar $s \in \K$ obtained by 
  dot product of the row vector $o$ with $v \in \K^{X \times 1}$.
  $A^*$ is the set of all words over $A$. $\epsilon$ is the empty word and $aw$ is the
  concatenation of a symbol $a$ to the word $w \in A^*$.
  A weighted language is a function $\psi: A^* \to \K$.
  A function mapping each state vector into its 
  accepted language, $\sqmap{\cdot}: \K^X \to K^{A^*}$ is defined as follows for every weighted automaton:

  \begin{equation*}
    \begin{aligned}
      \forall v \in \K^X, a \in A, w \in A^* \quad \quad
      \sqmap{v}(\epsilon) = o(v) \quad \quad
      \sqmap{v}(aw) = \sqmap{t_a(v)}(w)  
    \end{aligned}
  \end{equation*}

  Two vectors $v_1, v_2 \in \K^{X\times 1}$ are called language equivalent, denoted with 
  $v_1 \sim v_2 $ if and only if $ \sqmap
  {v_1} = \sqmap{v_2}$. One can extend the notion of language equivalence to states rather
  than for vectors by assigning to each state $x \in X$ the corresponding unit vector 
  $e_x \in \K^X$. When given an initial state $i$ for a weighted automaton, the language 
  of the automaton can be defined as $\sqmap{i}$.
\end{defn}

@@

% TODO linear weighted automata
\begin{defn}
  \textbf{} \\

@@

@@

\subsection{Forward Partition Refinement Algorithm for the Largest Weighted Bisimulation}
\label{sec:algo2}

The algorithm is defined by the iterative method:
\begin{eqnarray}
  R_0 = \mker{o}, & \quad & R_{i+1} = R_i \cap \bigcap_{a \in A} t_a(R_i)^{-1} \label{fwd} 
\end{eqnarray}
The algorithm stops when $R_{j+1} = R_j$. Proof is available in section 4.1 of \cite{BONCHI201277}

@@

@@

The algorithms and data structures for this paper are implemented in the Go programming 
language. This implementation makes use of the \textit{Gonum} library for numerical 
computations. We only import the Gonum libraries for matrices and linear algebra 
and visual plotting of samples and functions.
Real numbers are implemented with double precision floating point numbers,  
known as the \texttt{float64} type in the Go programming language.

@@

\subsection{Input file format}

\subsection{Calculating the intersection of linear subspaces}

To compute algorithm (\ref{fwd}), we need an efficient way to calculate a basis 
of the intersection of an arbitrary number of linear subspaces of $\R^n$, given their 
spanning sets.

\begin{defn}
  \textbf{Algorithm for a basis of the intersection of subspaces} \\
  Defined in file \texttt{lin/intersect.go}.
  Let's consider two linear subspaces of 
  $\R^n$ defined with their spanning sets of column vectors: 
  $U = \spanset{u_1, u_2, \hdots, u_p}$ 
  and $W = \spanset{w_1, w_2, \hdots, w_k}$.
  We create the block matrix $A \in \R^{n \times (p+k)}$ defined as follows:

  \begin{equation*}
    \begin{aligned}
      a_{ij} = \begin{cases}
        (u_j)_i & \text{for  } j = 1, \hdots, p \quad i = 1, \hdots, n \\
        (w_{j-p})_i & \text{for  } j = p+1, \hdots, p+k \quad i = 1, \hdots, n \\
      \end{cases}
    \end{aligned}
  \end{equation*}

  The matrix will have the form
  \begin{equation*}
    A = \begin{bmatrix}
      \vvvec{}{u_1}{\,}, \hdots \vvvec{}{u_p}{\,}, 
      \vvvec{}{w_1}{\,}, \hdots \vvvec{}{w_k}{\,} 
    \end{bmatrix}
  \end{equation*}

  It follows that $\mker{A}$ is a basis for $U \cap W$. This can be done with an 
  arbitrary finite number of subspaces of $\R^n$, as long as their spanning set is 
  \textit{finite}.
\end{defn}

@@

\begin{prop}
  Within finite dimensional spaces, the annihilator $W^0$ is isomorphic
  to the orthogonal complement.  
\end{prop}

\begin{proof}
  Let $V$ be an inner product space over the field $\K$ with an inner product
  $\langle \cdot , \cdot \rangle : V \times V \to \K$ defined. 
  With the inner product defined on $V$, every linear functional can be 
  represented with a vector: if $\xi : V \to \K$ is a functional, $c$ 
  is an unique vector such that $\forall x \in V \: . \: \langle c, x \rangle = \xi(x)$.
  The map associating every $\xi \in V^*$ with the vector $c \in V$ that corresponds 
  to $\xi$ with respect to the inner product is a linear isomorphism. Through this 
  isomorphism, the annihilator $W^0$ becomes $W^\perp$.
\end{proof}

>>>>

\theoremstyle{plain}% default
\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newtheorem*{cor}{Corollary}
\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{conj}{Conjecture}[section]
\newtheorem{exmp}{Example}[section]
\newtheorem{exrc}[exmp]{Exercise}
\theoremstyle{remark}
\newtheorem*{comm}{Comment}
\newtheorem*{note}{Note}
\newtheorem{case}{Case}

@@

\newcommand{\diag}[1]{\text{diag}\left(#1\right)}
\newcommand{\mspan}[1]{\text{span}\left(#1\right)}
\newcommand{\dom}{\text{dom}}
\newcommand{\mker}[1]{\text{ker}\left(#1\right)}
\newcommand{\mrank}[1]{\text{rank}\left(#1\right)}
\newcommand{\llwb}{\approx_{\mathcal{L}}}
\newcommand{\langlwa}[2]{\sqmap{#1}^\mathcal{L}_{#2}}

@@

\author{
  Alessandro Cheli \\
  Undergraduate Student \\
  Department of Computer Science \\
  Università di Pisa\\
  Pisa, PI 56127 \\
  \texttt{a.cheli6@studenti.unipi.it} \\
  %% examples of more authors
  %% \And
  %% Elias D.~Striatum \\
  %% Department of Electrical Engineering\\
  %% Mount-Sheikh University\\
  %% Santa Narimana, Levand \\
  %% \texttt{stariate@ee.mount-sheikh.edu} \\
}

@@

In \cite{DBLP:journals/corr/Bonchi0K17}, up-to techniques are defined for
weighted systems over arbitrary semirings, while in \cite{BONCHI201277}, up-to techniques
are defined for Linear Weighted Automata (LWAs), under a more abstract coalgebraic perspective.

@@

\begin{note}
  Given two vector spaces $V_1, V_2$ we write $V_1 + V_2$ to denote $\mspan{V_1 \cup V_2}$
\end{note}

\begin{defn}
  A \textit{weighted automaton} over a field $\K$ and an alphabet $A$ is a triple 
  $(X,o,t)$ such that $X$ is a finite set of states, 
  $t = \left\lbrace t_a : X \to \K\right\rbrace_{a \in A}$
  is a set of transition functions indexed over the symbols of the alphabet $A$ and $o : X 
  \to \K$ is the output function. 
  $A^*$ is the set of all words over $A$. $\epsilon$ is the empty word and $aw$ is the
  concatenation of a symbol $a$ to the word $w \in A^*$.
  A weighted language is a function $\psi: A^* \to \K$.
  A function mapping each state vector into its 
  accepted language, $\sqmap{\cdot}: \K^X \to \K^{A^*}$ is defined as follows for every weighted automaton:

  \begin{equation*}
    \begin{aligned}
      \forall v \in \K^X, a \in A, w \in A^* \quad \quad
      \sqmap{v}(\epsilon) = o(v) \quad \quad
      \sqmap{v}(aw) = \sqmap{t_a(v)}(w)  
    \end{aligned}
  \end{equation*}
\end{defn}

Two vectors $v_1, v_2 \in \K^{X\times 1}$ are weighted called language equivalent, denoted with 
$v_1 \sim_l v_2 $ if and only if $ \sqmap
{v_1} = \sqmap{v_2}$. One can extend the notion of language equivalence to states rather
than for vectors by assigning to each state $x \in X$ the corresponding unit vector 
$e_x \in \K^X$. When given an initial state $i$ for a weighted automaton, the language 
of the automaton can be defined as $\sqmap{i}$.

@@

%===================================================================================

We omit the coalgebraic definition for \textit{linear weighted automata} seen in 
\cite{BONCHI201277} and give a more intuitive definition. 
In this implementation, we focus only on weighted automata defined over 
the field of real numbers $\R$.

\begin{defn}
  A \textit{linear weighted automaton} (in short, LWA) over the field $\K$ and an alphabet $A$
  is a triple  $L = (V, o, \left\lbrace t_a \right\rbrace_{a \in A})$ 
  where $V$ is a vector space representing the state space, 
  $o: V \to \K$ is a linear map associating to each state its output weight,
  and $t = \left\lbrace t_a = V \times V \right\rbrace_{a \in A}$ is
  the set of transition functions, represented with liner maps 
  that for each input $a \in A$ associate the next state, in this case a vector
  in $V$.
  As in \cite{boreale2009weighted}, we have that $\dim{(L)} = \dim{(V)}$.
\end{defn}

Given a weighted automaton, one can build an isomorphic linear weighted automaton
by considering the free vector space generated by the set of states $X$ in the WA,
and by linearizing $o$ and $t$. If $X$ is finite, we can use the same matrices for 
$t$ and $o$ in both the WA and the corresponding LWA.
We are only considering a finite number of states and therefore finite dimensional
vector spaces. Let $n$ be the (finite) number of states in an WA.
We have that in the corresponding LWA, the transition functions $t_a$ are still
 represented as
$\K^{n \times n}$ matrices. $o \in \K^{1 \times n}$ is represented as a row vector.
$t_a(v)$ denotes the vector obtained by multiplying the matrix $t_a$ by the column 
vector $v  \in \K^{n \times 1}$. $o(v)$ denotes the scalar $s \in \K$ obtained by 
dot product of the row vector $o$ with $v \in \K^{n \times 1}$.

\begin{defn}
  The language recognized by a vector $v \in V$ of an LWA $(V,o,t)$ is defined
  for all words $w \in A*$ as $\langlwa{v}{V}(w) = o(v_n)$ where $v_n$ is the 
  vector reached from $v$ through the composition of the transition functions
  corresponding to the words in $w$.

  \begin{equation*}
    \begin{aligned}
      \langlwa{v}{V}(w) = \begin{cases}
        o(v) & \text{if } w = \epsilon \\ 
        \langlwa{t_a(v)}{V}(w') & \text{if } w = aw' 
      \end{cases}
    \end{aligned}
  \end{equation*}

@@

We define $\llwb$ as the behavioral equivalence for a given LWA $(V, o, t)$ as

\begin{equation}
  \forall v_1, v_2 \in V, \; v_1 \llwb v_2 \iff \langlwa{v_1}{V} = \langlwa{v_2}{V}
\end{equation}

\begin{lem}
  $\llwb$ coincides with $\sim_l$:

  Let $(X, o, t)$ be a WA and $(\K^X, o^\sharp, t^\sharp)$ the corresponding linear 
  weighted automaton. Then $\forall x \in X, \;\; \sqmap{x} = \langlwa{x}{\K^X}$
\end{lem}

\begin{proof}
  Proved in section 3.2 of \cite{BONCHI201277}
\end{proof}

%=========================================================================

\section{The Problem}
In this work, we

@@

We now recall the backwards algorithm for computing $\llwb$ defined in \cite{BONCHI201277}.

@@

%\subsection{Forward Partition Refinement Algorithm for the Largest Weighted Bisimulation}
%\label{sec:algo2}
%
%The algorithm is defined by the iterative method:
%\begin{eqnarray}
%  R_0 = \mker{o}, & \quad & R_{i+1} = R_i \cap \bigcap_{a \in A} t_a(R_i)^{-1} \label{fwd} 
%\end{eqnarray}
%The algorithm stops when $R_{j+1} = R_j$. Proof is available in section 4.1 of \cite{BONCHI201277}

@@

The algorithms and data structures for this paper are implemented in the Go programming 
language. This implementation makes use of the Gonum library for numerical 
computations. We only import the Gonum libraries for matrices and linear algebra 
and visual plotting of samples and functions.
Real numbers are implemented with double precision floating point numbers,  
known as the \texttt{float64} type in the Go programming language.

@@

@@

\begin{prop}
  If $V$ is a finite dimensional vector space defined with an inner product
  $\langle \cdot , \cdot \rangle$ and $W$ is a subspace of $V$
  then the image of the annhilitaor $W^0$ through the linear 
  isomorphism $\varphi: V^* \to V$ induced by the inner product, 
  is the orthogonal of $W$ with respect to the said inner product.
\end{prop}

\begin{proof}
  Let $V$ be an inner product space over the field $\K$ with an inner product defined as
  $\langle \cdot , \cdot \rangle : V \times V \to \K$. 
  Every linear functional can be 
  represented with a vector. Let $\xi : V \to \K$ be a functional, 
  $\xi \in  W^0$. Because $\xi(w)=0 \quad \forall  w \in W$, 
  if $v$ represents $\xi$ we have that $(v, w)=\xi(w)=0$ for all $w \in W$. 
  We obtain that $\varphi(W^0) \subseteq W^{\perp}$.
  If $v \in W^\perp$  
  then the functional $x \mapsto (v, x)$ cancels over $W$ 
  (by the definition of orthogonality).
\end{proof}
